---
title: "Biostat 203B Homework 1 - Corrections"
author: Danielle LaVine
subtitle: Due Jan 21 @ 11:59PM
output: html_document
---

<span style="color: #CC0033;">Corrections completed: Q3.(4,5,6). Fixed reproducibility problems by changing to EVAL = T. Adjusted line spacing to fit the 80-character limit. </span>

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Display machine information for reproducibility:
```{r, eval=F}
sessionInfo()
```

## Q1. Git/GitHub

**No handwritten homework reports are accepted for this course.** We work with Git and GitHub. Efficient and abundant use of Git, e.g., frequent and well-documented commits, is an important criterion for grading your homework.

1. Apply for the [Student Developer Pack](https://education.github.com/pack) at GitHub using your UCLA email. You'll get GitHub Pro account for free (unlimited public and private repositories).

2. Create a **private** repository `biostat-203b-2022-winter` and add `Hua-Zhou` and `maschepps` as your collaborators with write permission.

3. Top directories of the repository should be `hw1`, `hw2`, ... Maintain two branches `main` and `develop`. The `develop` branch will be your main playground, the place where you develop solution (code) to homework problems and write up report. The `main` branch will be your presentation area. Submit your homework files (R markdown file `Rmd`, `html` file converted from R markdown, all code and extra data sets to reproduce results) in `main` branch.

4. After each homework due date, teaching assistant and instructor will check out your main branch for grading. Tag each of your homework submissions with tag names `hw1`, `hw2`, ... Tagging time will be used as your submission time. That means if you tag your `hw1` submission after deadline, penalty points will be deducted for late submission.

5. After this course, you can make this repository public and use it to demonstrate your skill sets on job market.

## Q2. Data ethics training

This exercise (and later in this course) uses the [MIMIC-IV data](https://mimic-iv.mit.edu), a freely accessible critical care database developed by the MIT Lab for Computational Physiology. Follow the instructions at <https://mimic.mit.edu/docs/gettingstarted/> to (1) complete the CITI `Data or Specimens Only Research` course and (2) obtain the PhysioNet credential for using the MIMIC-IV data. Display the verification links to your completion report and completion certificate here. (Hint: The CITI training takes a couple hours and the PhysioNet credentialing takes a couple days; do not leave it to the last minute.)

**Solution: **

Completion Report Link : https://www.citiprogram.org/verify/?k2eaee88d-e070-482b-ad6b-048aa9102214-46575325

Completion Certificate Link: https://www.citiprogram.org/verify/?we6a0240e-a0c1-449f-901f-844f99866ced-46575325

## Q3. Linux Shell Commandshttps://mimic.mit.edu/docs/iv/

1. The `/mnt/mimiciv/1.0` folder on teaching server contains data sets from MIMIC-IV. Refer to the documentation <> for details of data files.  
    ```{bash}
    ls -l /mnt/mimiciv/1.0
    ```
Please, do **not** put these data files into Git; they are big. Do **not** copy them into your directory. Do **not** decompress the gz data files. These create unnecessary big files on storage and are not big data friendly practices. Just read from the data folder `/mnt/mimiciv/1.0` directly in following exercises. 

    Use Bash commands to answer following questions.

2. Display the contents in the folders `core`, `hosp`, `icu`. Why are these data files distributed as `.csv.gz` files instead of `.csv` (comma separated values) files? Read the page <https://mimic.mit.edu/docs/iv/> to understand what's in each folder.

**Solution: **
These data files are distributed as `.csv.gz` files instead of `.csv` files because
they are compressed into a smaller file size, and will take up less room wherever 
those files are stored. The following lines of code display the contents in the folders
`core`, `hosp`, and `icu`:
```{bash}
ls -l /mnt/mimiciv/1.0/core
```

```{bash}
ls -l /mnt/mimiciv/1.0/hosp
```

```{bash}
ls -l /mnt/mimiciv/1.0/icu
```


3. Briefly describe what bash commands `zcat`, `zless`, `zmore`, and `zgrep` do.

**Solution: **
`zcat` uncompressed compressed files, such as those in part 2. `zless` and `zmore`
are the anologs to the `less` and `more` functions, but work for compressed files. 
`zless` and `zmore` allow users to read text files a page at a time. `zgrep` allows 
users to use `grep` on compressed files. Information was obtained by the following: 

```{bash, eval=F}
man zcat
```
```{bash, eval=F}
man zless
```

```{bash, eval=F}
man zmore
```

```{bash, eval=F}
man zgrep
```



4. What's the output of following bash script?
    ```{bash, eval=F}
    for datafile in /mnt/mimiciv/1.0/core/*.gz
      do
        ls -l $datafile
      done
    ```

Display the number of lines in each data file using a similar loop.

**Solution: **
The above bash script lists the files in the core folder which are in gz file 
format. The following displays the number of lines in each data file:

```{bash}
for datafile in  /mnt/mimiciv/1.0/core/*.gz
  do 
     zcat $datafile | awk 'END {print NR}' 
  done
```

5. Display the first few lines of `admissions.csv.gz`. How many rows are in this data file? How many unique patients (identified by `subject_id`) are in this data file? (Hint: combine Linux commands `zcat`, `head`/`tail`, `awk`, `sort`, `uniq`, `wc`, and so on.)

**Solution: **
```{bash}
zcat /mnt/mimiciv/1.0/core/admissions.csv.gz | head -5
```

```{bash}
zcat /mnt/mimiciv/1.0/core/admissions.csv.gz | awk 'END {print NR}' 
``` 
```{bash}
zcat /mnt/mimiciv/1.0/core/admissions.csv.gz | awk -F, '{print $1}' | sort | 
uniq|
awk 'END {print NR}'
```
There are 523741 total rows in this data file including the header, and there are 356878 unique patients. 

6. What are the possible values taken by each of the variable `admission_type`, `admission_location`, `insurance`, and `ethnicity`? Also report the count for each unique value of these variables. (Hint: combine Linux commands `zcat`, `head`/`tail`, `awk`, `uniq -c`, `wc`, and so on.)


**Solution: **
Show the possible values of `admission_type`
```{bash}
zcat /mnt/mimiciv/1.0/core/admissions.csv.gz |  awk -F, '{print $6}' |sort |
uniq -c
``` 
Print the number of unique values of `admission_type`
```{bash}
zcat /mnt/mimiciv/1.0/core/admissions.csv.gz |  awk -F, '{print $6}' |sort |
uniq -c|
awk 'END {print NR}'
```


Show the possible values of `admission_location`
```{bash}
zcat /mnt/mimiciv/1.0/core/admissions.csv.gz |  awk -F, '{print $7}' |sort |
uniq -c
``` 
Print the number of unique values of `admission_location`
```{bash}
zcat /mnt/mimiciv/1.0/core/admissions.csv.gz |  awk -F, '{print $7}' |sort |
uniq -c | awk 'END {print NR}'
```


Show the possible values of `insurance`
```{bash}
zcat /mnt/mimiciv/1.0/core/admissions.csv.gz |  awk -F, '{print $9}' |sort |
uniq -c
``` 
Print the number of unique values of `insurance`
```{bash}
zcat /mnt/mimiciv/1.0/core/admissions.csv.gz |  awk -F, '{print $9}' |sort |
uniq -c | awk 'END {print NR}'
```
Show the possible values of `ethnicity`
```{bash}
zcat /mnt/mimiciv/1.0/core/admissions.csv.gz |  awk -F, '{print $12}' |sort |
uniq -c
``` 
Print the number of unique values of `admission_type`
```{bash}
zcat /mnt/mimiciv/1.0/core/admissions.csv.gz |  awk -F, '{print $12}' |sort |
uniq -c | awk 'END {print NR}'
```
## Q4. Who's popular in Price and Prejudice

1. You and your friend just have finished reading *Pride and Prejudice* by Jane Austen. Among the four main characters in the book, Elizabeth, Jane, Lydia, and Darcy, your friend thinks that Darcy was the most mentioned. You, however, are certain it was Elizabeth. Obtain the full text of the novel from <http://www.gutenberg.org/cache/epub/42671/pg42671.txt> and save to your local folder. 
    ```{bash}
    wget -nc http://www.gutenberg.org/cache/epub/42671/pg42671.txt
    ```
Explain what `wget -nc` does. Do **not** put this text file `pg42671.txt` in Git. Complete the following loop to tabulate the number of times each of the four characters is mentioned using Linux commands.

**Solution: **


`wget` follows the link provided and creates a copy of the content on the website. 
The `-nc` option ensures that if the file is downloaded several times, that the file is overwritten. 


    ```{bash}
    wget -nc http://www.gutenberg.org/cache/epub/42671/pg42671.txt
    for char in Elizabeth Jane Lydia Darcy
    do
      echo $char:
     grep $char pg42671.txt| wc -l
    done
    ```


2. What's the difference between the following two commands?
    ```{bash}
    echo 'hello, world' > test1.txt
    ```
    and
    ```{bash eval=FALSE}
    echo 'hello, world' >> test2.txt
    ```

**Solution: **
`>` creates a new text document containing the words "hello, world", whereas
`>>` creates a blank text document, and adds the words "hello, world". 

3. Using your favorite text editor (e.g., `vi`), type the following and save the file as `middle.sh`:
    ```{bash, eval = FALSE}
    #!/bin/sh
    # Select lines from the middle of a file.
    # Usage: bash middle.sh filename end_line num_lines
    head -n "$2" "$1" | tail -n "$3"
    ```
    
    
Using `chmod` make the file executable by the owner, and run 


Explain the output. Explain the meaning of `"$1"`, `"$2"`, and `"$3"` in this shell script. Why do we need the first line of the shell script?

**Solution: **
```{bash}
chmod u+x ./middle.sh
```

    ```{bash eval=FALSE}
    ./middle.sh ~/biostat-203b-2022-winter/pg42671.txt 20 5
    ```
The output of this bash command returns 5 lines of the text file ending at line 20. 
The symbols `"$1"`, `"$2"` and `"$3"` indicate where to place which arguments of the function in the shell script. `"$1"` in this case represents `~/biostat-203b-2022-winter/pg42671.txt`, `"$2"` is `20` and `"$3"` is '5'. 


### Q5. More fun with Linux

Try following commands in Bash and interpret the results: `cal`, `cal 2021`, `cal 9 1752` (anything unusual?), `date`, `hostname`, `arch`, `uname -a`, `uptime`, `who am i`, `who`, `w`, `id`, `last | head`, `echo {con,pre}{sent,fer}{s,ed}`, `time sleep 5`, `history | tail`.


**Solution: **
The `cal` command outputs the calendar for the current month. `cal 2021` outputs 
the entire 12-month calendar for 2021. `cal 9 1752` outputs the calendar for September
1752, however it appears that it omits dates between September 2 and September 14. 
`date` gives the current day of the week, calendar day, time and year. 

```{bash}
  cal
  cal 2021
  cal 9 1752
  date
```

`hostname` gives the name of the server. `arch` prints the machine name (info from `man arch`), `uname -a` prints all system information. 

```{bash}
  hostname 
  arch
  uname -a
```

`uptime` displays how long the server has been running, the number of users currently
logged in, and the load averages for three time periods. 
```{bash}
uptime
```
`whoami` displays my user name. 
```{bash}
whoami
```
`who` shows who is currently logged onto the server. 
```{bash}
who
```
From `man w`: the `w` command shows how many people are logged on, how long it's been running, and the load averages for three time periods. This command also outputs the username, login time, the time used by the user for each user. 
```{bash}
w
```
`id` prints the user id and group id for the user. 
```{bash}
id
```
The command `last | head` lists the last 10 users who logged into the server. 
```{bash}
last | head 
```
The command `echo {con,pre}{sent,fer}{s,ed}` concatenates each combination of the two
prefixes and suffixes listed in the three braces, in order of how they are listed. 
```{bash}
echo {con,pre}{sent,fer}{s,ed}
```
The command `time sleep 5` makes the computer pause for 5 seconds. 
```{bash}
time sleep 5
```
 `history | tail` displays the last 10 lines of the command history. 
```{bash}
history | tail
```