---
title: "hw2_sol_corrections"
author: "Danielle LaVine"
date: "1/25/2022"
output: html_document
---

<span style="color: #CC0033;">Corrections completed: Q5 part 3, Q6, and Q7. Also fixed issues with the 80 character limit. </span>

Display machine information for reproducibility:
```{r}
sessionInfo()
```

```{r setup, message=F}
knitr::opts_chunk$set(echo = TRUE, cache = TRUE, cache.lazy = FALSE)
library(tidyverse)
library(data.table)
library(lubridate)
```

```{r}
os <- sessionInfo()$running
if (str_detect(os, "Linux")) {
  mimic_path <- "/mnt/mimiciv/1.0"
} else if (str_detect(os, "macOS")) {
  mimic_path <- "/Users/huazhou/Documents/Box Sync/MIMIC/mimic-iv-1.0"
}


```

In this exercise, we use tidyverse (ggpot2, dplyr, etc) to explore the [MIMIC-IV](https://mimic.mit.edu/docs/iv/) data introduced in [homework 1](https://ucla-biostat-203b.github.io/2022winter/hw/hw1/hw1.html) and to build a cohort of ICU stays.

```{r}
# tree -s -L 2 /Users/huazhou/Documents/Box\ Sync/MIMIC/mimic-iv-1.0
system(str_c("tree -s -L 2 ", shQuote(mimic_path)), intern = TRUE)
```

## Q1. `read.csv` (base R) vs `read_csv` (tidyverse) vs `fread` (data.table)

There are quite a few utilities in R for reading plain text data files. Let us test the speed of reading a moderate sized compressed csv file, `admissions.csv.gz`, by three programs: `read.csv` in base R, `read_csv` in tidyverse, and `fread` in the popular data.table package. 

Which function is fastest? Is there difference in the (default) parsed data types? (Hint: R function `system.time` measures run times.)

For later questions, we stick to the tidyverse.

**Solution: **
The fastest of the three functions is the `fread` function, with an elapsed time 
of .921 seconds, versus the `read_csv` function which took 2.311 seconds, and the 
`read.csv` function whicb took 45.674 seconds. 
```{r}
system.time(read.csv(str_c(mimic_path, '/core/admissions.csv.gz')))
```
```{r}
system.time(read_csv(str_c(mimic_path, '/core/admissions.csv.gz')))
```
```{r}
system.time(fread(str_c(mimic_path, '/core/admissions.csv.gz')))
```

## Q2. ICU stays

`icustays.csv.gz` (<https://mimic.mit.edu/docs/iv/modules/icu/icustays/>) contains data about Intensive Care Units (ICU) stays. The first 10 lines are
```{r}
system(
  str_c(
    "zcat < ", 
    shQuote(str_c(mimic_path, "/icu/icustays.csv.gz")), 
    " | head"
    ), 
  intern = TRUE
)
```

1. Import `icustatys.csv.gz` as a tibble `icustays_tble`. 

2. How many unique `subject_id`? Can a `subject_id` have multiple ICU stays? 

3. For each `subject_id`, let's only keep the first ICU stay in the tibble `icustays_tble`.


**Solution: **
1. 
```{r}
icustays_tble <- read.csv(str_c(mimic_path, "/icu/icustays.csv.gz"))
```

```{r}
head(icustays_tble, length=8)
```

2. There are 53150 unique `subject_id`, and some have stayed in the ICU multiple times.
```{r}
icustays_tble %>% group_by(subject_id) %>% summarise(numstays = length(subject_id))
```

3. 
```{r}
icustays_tble <- icustays_tble %>% group_by(subject_id) %>%
  slice_min(n = 1, order_by = as.numeric(intime))

head(icustays_tble)
```

## Q3. `admission` data

Information of the patients admitted into hospital is available in `admissions.csv.gz`. See <https://mimic.mit.edu/docs/iv/modules/core/admissions/> for details of each field in this file. The first 10 lines are
```{r}
system(
  str_c(
    "zcat < ", 
    shQuote(str_c(mimic_path, "/core/admissions.csv.gz")), 
    " | head"
    ), 
  intern = TRUE
)
```

1. Import `admissions.csv.gz` as a tibble `admissions_tble`.


2. Let's only keep the admissions that have a match in `icustays_tble` according to `subject_id` and `hadmi_id`.

3. Summarize the following variables by graphics. 

- admission year  
- admission month  
- admission month day  
- admission week day  
- admission hour (anything unusual?)  


**Solution: **

1.
```{r}
admissions_table <- read_csv(str_c(mimic_path, "/core/admissions.csv.gz"))
```

2. 

```{r}
admissions_table <- semi_join(admissions_table, icustays_tble, 
                              by = c("subject_id", "hadm_id"))
```


3. 
```{r}
#Admissions by year
ggplot(data = admissions_table, aes(x= year(admittime)))+ geom_bar() +
  ggtitle("Summary of Admissions by Year") + xlab("Year") + theme_minimal()

#Admissions by month
ggplot(data = admissions_table, aes(x= month(admittime, label=TRUE)))+
  geom_bar() + ggtitle( "Summary of Admissions by Month")+xlab("Month") +
  theme_minimal(); 

#Admissions by day of the month
ggplot(data = admissions_table, aes(x= mday(admittime))) + geom_bar() +
  ggtitle("Summary of Admissions by Day of the Month") + 
  xlab("Day of the Month")+theme_minimal()

#Admissions by day of the week
ggplot(data = admissions_table, aes(x= wday(admittime, label=T))) +
  geom_bar() + ggtitle("Summary of Admissions by Day of the Week") + 
  xlab("Day of the Week") + 
  theme_minimal()

#Admissions by hour
ggplot(data = admissions_table, aes(x= hour(admittime))) + geom_bar() +
  ggtitle("Summary of Admissions by Hour of the Day") + xlab("Hour") + 
  theme_minimal()

```

There appears to be many admissions at midnight. We would probably expect that the number of admissions from 11 pm to 12 am and 12 am to 1 pm would not be so starkly different, but it might be due to input error. 

## Q4. `patients` data

Patient information is available in `patients.csv.gz`. See <https://mimic.mit.edu/docs/iv/modules/core/patients/> for details of each field in this file. The first 10 lines are
```{r}
system(
  str_c(
    "zcat < ", 
    shQuote(str_c(mimic_path, "/core/patients.csv.gz")), 
    " | head"
    ), 
  intern = TRUE
)
```

1. Import `patients.csv.gz` (<https://mimic.mit.edu/docs/iv/modules/core/patients/>) as a tibble `patients_tble` and only keep the patients who have a match in `icustays_tble` (according to `subject_id`).



2. Summarize variables `gender` and `anchor_age`, and explain any patterns you see.



**Solution: **
1. 

```{r}
patients_tble <- read_csv(str_c(mimic_path, "/core/patients.csv.gz")) %>%
  semi_join(icustays_tble, by = "subject_id")
```

2.

```{r}
patients_tble %>% summarise(n_male = sum(gender=="M"), 
                            n_female = sum(gender =="F"), 
                            avg_anchor_age = mean(anchor_age), 
                            avg_M_age = mean(anchor_age[gender == "M"]),
                            avg_F_age = mean(anchor_age[gender == "F"]))
```


## Q5. Lab results

`labevents.csv.gz` (<https://mimic.mit.edu/docs/iv/modules/hosp/labevents/>) contains all laboratory measurements for patients. The first 10 lines are
```{r}
system(
  str_c(
    "zcat < ", 
    shQuote(str_c(mimic_path, "/hosp/labevents_filtered_itemid.csv.gz")), 
    " | head"
    ), 
  intern = TRUE
)
```
`d_labitems.csv.gz` is the dictionary of lab measurements. 
```{r}
system(
  str_c(
    "zcat < ", 
    shQuote(str_c(mimic_path, "/hosp/d_labitems.csv.gz")), 
    " | head"
    ), 
  intern = TRUE
)
```

1. Find how many rows are in `labevents.csv.gz`.

2. We are interested in the lab measurements of creatinine (50912), potassium (50971), sodium (50983), chloride (50902), bicarbonate (50882), hematocrit (51221), white blood cell count (51301), glucose (50931), magnesium (50960), and calcium (50893). Retrieve a subset of `labevents.csv.gz` only containing these items for the patients in `icustays_tble` as a tibble `labevents_tble`. 

    Hint: `labevents.csv.gz` is a data file too big to be read in by the `read_csv` function in its default setting. Utilize the `col_select` and `lazy` options in the `read_csv` function to reduce the memory burden.

3. Further restrict `labevents_tble` to the first lab measurement during the ICU stay. 
4. Summarize the lab measurements by appropriate numerics and graphics. 


**Solution: **
Part 1:
```{r}
lab_events_tble <- read_csv(str_c(
  mimic_path, "/hosp/labevents_filtered_itemid.csv.gz"))
```

```{r}
num_rows <- dim(lab_events_tble)[1]; num_rows
```

There are 30198826 rows in `labevents.csv.gz`.

Part 2: 
```{r}
labevents_tble <- lab_events_tble %>%
  filter(itemid %in% c(50912, 50971, 50983, 50902, 50882, 51221, 
                       51301, 50931, 50960, 50893))
```


Part 3:  
```{r}
labevents_tble <- labevents_tble %>%  group_by(subject_id, itemid) %>% 
  slice_min(n = 1, order_by = as.numeric(charttime))
```



Part 4: 
```{r}
lab_meas <- labevents_tble %>%
 pivot_wider(names_from = itemid, values_from = valuenum, values_fn = {min})
colnames(lab_meas)[3:12] <- c("hematocrit", "wbc_count", "bicarbonate", 
                              "calcium", "chloride", "creatinine", "magnesium",
                              "potassium", "sodium", "glucose")
```

```{r}
labevents_tble %>% group_by(itemid) %>% summarise(mean = 
                                                    mean(na.omit(valuenum)),
                                                  sd = sd(na.omit(valuenum)), 
                                                  min = min(na.omit(valuenum)),
                                                  max = max(na.omit(valuenum)))
```

```{r}
ggplot(lab_meas, aes(x=hematocrit))+ geom_histogram()

ggplot(lab_meas, aes(x=wbc_count))+ geom_histogram()

ggplot(lab_meas, aes(x=bicarbonate))+ geom_histogram()

ggplot(lab_meas, aes(x=calcium))+ geom_histogram()

ggplot(lab_meas, aes(x=chloride))+ geom_histogram()

ggplot(lab_meas, aes(x=creatinine))+ geom_histogram()

ggplot(lab_meas, aes(x=magnesium))+ geom_histogram()

ggplot(lab_meas, aes(x=potassium))+ geom_histogram()

ggplot(lab_meas, aes(x=sodium))+ geom_histogram()

ggplot(lab_meas, aes(x=glucose))+ geom_histogram()
```


## Q6. Vitals from charted events

`chartevents.csv.gz` (<https://mimic.mit.edu/docs/iv/modules/icu/chartevents/>) contains all the charted data available for a patient. During their ICU stay, the primary repository of a patient’s information is their electronic chart. The `itemid` variable indicates a single measurement type in the database. The `value` variable is the value measured for `itemid`. The first 10 lines of `chartevents.csv.gz` are
```{r}
system(
  str_c(
    "zcat < ", 
    shQuote(str_c(mimic_path, "/icu/chartevents_filtered_itemid.csv.gz")), 
    " | head"), 
  intern = TRUE
)
```
`d_items.csv.gz` (<https://mimic.mit.edu/docs/iv/modules/icu/d_items/>) is the dictionary for the `itemid` in `chartevents.csv.gz`. 
```{r}
system(
  str_c(
    "zcat < ", 
    shQuote(str_c(mimic_path, "/icu/d_items.csv.gz")), 
    " | head"), 
  intern = TRUE
)
```

1. We are interested in the vitals for ICU patients: heart rate (220045), mean non-invasive blood pressure (220181), systolic non-invasive blood pressure (220179), body temperature in Fahrenheit (223761), and respiratory rate (220210). Retrieve a subset of `chartevents.csv.gz` only containing these items for the patients in `icustays_tble` as a tibble `chartevents_tble`.

2. Further restrict `chartevents_tble` to the first vital measurement during the ICU stay. 

3. Summarize these vital measurements by appropriate numerics and graphics. 


**Solution: **
1. 
```{r}
chartevents_tble <- read_csv(str_c(mimic_path, "/icu/chartevents_filtered_itemid.csv.gz")) %>% 
  filter(itemid %in% c(220045, 220181, 220179, 223761, 220210))
```




2. 
```{r}
chartevents_tble <- chartevents_tble %>% group_by(subject_id, itemid) %>% 
  slice_min(n = 1, order_by = as.numeric(charttime))
```





```{r}
vitals <- chartevents_tble %>% 
  pivot_wider(names_from = itemid, values_from = valuenum)
colnames(vitals)[5:9] <- c("temp_far", "resp_rate", "heart_rate", "syst_bp", 
                           "mean_bp" )
```





3.
```{r}
ggplot(vitals, aes(x=resp_rate)) + 
  geom_histogram()+ggtitle("Distribution of Respiratory Rate")+
  theme_minimal()

ggplot(vitals, aes(x=heart_rate)) + 
  geom_histogram()+ggtitle("Distribution of Heart Rate")+
  theme_minimal()

ggplot(vitals, aes(x=mean_bp)) + 
  geom_histogram()+ggtitle("Distribution of Mean Blood Pressure")+
  theme_minimal()

ggplot(vitals, aes(x=syst_bp)) + 
  geom_histogram()+ggtitle("Distribution of Systolic Blood Pressure")+
  theme_minimal()

ggplot(vitals, aes(x=temp_far)) + 
  geom_histogram()+ggtitle("Distribution of Temperature in Fahrenheit")+
  theme_minimal()
```



## Q7. Putting things together

Let us create a tibble `mimic_icu_cohort` for all ICU stays, where rows are  

- first ICU stay of each unique adult (age at admission > 18)

and columns contain at least following variables  

- all variables in `icustays.csv.gz`  
- all variables in `admission.csv.gz`  
- all variables in `patients.csv.gz`  
- first lab measurements during ICU stay  
- first vital measurements during ICU stay
- an indicator variable `thirty_day_mort` whether the patient died within 30 days of hospital admission (30 day mortality)


**Solution: **


```{r}
#all variables in the three tables
mimic_icu_cohort <- icustays_tble %>%
  left_join(admissions_table, by = c("subject_id", "hadm_id"))  %>% 
  left_join(patients_tble, by = "subject_id") %>%
  filter(anchor_age > 18) 


head(mimic_icu_cohort)
  
```


```{r}
#adding first lab measurements
mimic_icu_cohort <- mimic_icu_cohort %>%
  inner_join(lab_meas, by = "subject_id") %>%
  inner_join(vitals, by = "subject_id")

head(mimic_icu_cohort)
```
```{r}
#adding the 30-day mortality column 
mimic_icu_cohort$thirty_day_mort <- 
  as.numeric((ymd_hms(mimic_icu_cohort$deathtime) - 
                ymd_hms(mimic_icu_cohort$admittime)) <= 30*60*24)
```


## Q8. Exploratory data analysis (EDA)

Summarize following information using appropriate numerics or graphs.

- `thirty_day_mort` vs demographic variables (ethnicity, language, insurance, marital_status, gender, age at hospital admission)

- `thirty_day_mort` vs first lab measurements

- `thirty_day_mort` vs first vital measurements

- `thirty_day_mort` vs first ICU unit

**Solution: **

```{r}
mimic_icu_cohort %>% 
  group_by(ethnicity) %>% 
  summarise(numdeaths = sum(na.omit(thirty_day_mort)))

ggplot(mimic_icu_cohort, aes(x = ethnicity, y = thirty_day_mort)) + 
  geom_col() + ggtitle("Thirty Day Mortality by Ethnicity") + 
  theme_minimal()
```
```{r}
mimic_icu_cohort %>% 
  group_by(language) %>% 
  summarise(numdeaths = sum(na.omit(thirty_day_mort)))

ggplot(mimic_icu_cohort, aes(x = language, y = thirty_day_mort)) + 
  geom_col() + ggtitle("Thirty Day Mortality by Language") + 
  theme_minimal()
```

```{r}
mimic_icu_cohort %>% 
  group_by(insurance) %>%
  summarise(numdeaths = sum(na.omit(thirty_day_mort)))

ggplot(mimic_icu_cohort, aes(x = insurance, y = thirty_day_mort)) + 
  geom_col() + ggtitle("Thirty Day Mortality by Insurance Type") + 
  theme_minimal()
```
```{r}
mimic_icu_cohort %>%
  group_by(marital_status) %>%
  summarise(numdeaths = sum(na.omit(thirty_day_mort)))

ggplot(mimic_icu_cohort, aes(x = marital_status, y = thirty_day_mort)) + 
  geom_col() + ggtitle("Thirty Day Mortality by Marital Status") + 
  theme_minimal()
```
```{r}
mimic_icu_cohort %>%
  group_by(gender) %>%
  summarise(numdeaths = sum(na.omit(thirty_day_mort)))

ggplot(mimic_icu_cohort, aes(x = gender, y= thirty_day_mort)) + geom_col() + 
  ggtitle("Thirty Day Mortality by Gender") + theme_minimal()
```
```{r}
mimic_icu_cohort %>%
  group_by(anchor_age) %>%
  summarise(numdeaths = sum(na.omit(thirty_day_mort)))

ggplot(mimic_icu_cohort, aes(x = anchor_age, y = thirty_day_mort)) + 
  geom_col() + ggtitle("Thirty Day Mortality by Age") + 
  theme_minimal()
```


```{r}
#Table summarizing the average first lab measurements by thirty day mortality.
mimic_icu_cohort %>% 
  group_by(thirty_day_mort) %>%
  summarise(mean_hematocrit = mean(na.omit(hematocrit)), 
            mean_wbc_count = mean(na.omit(wbc_count)), 
            mean_bicarbonate = mean(na.omit(bicarbonate)), 
            mean_calcium = mean(na.omit(calcium)),
            mean_chloride = mean(na.omit(chloride)), 
            mean_creatinine = mean(na.omit(creatinine)),
            mean_magnesium = mean(na.omit(magnesium)), 
            mean_potassium = mean(na.omit(potassium)),
            mean_sodium = mean(na.omit(sodium)), 
            mean_glucose = mean(na.omit(glucose))
  )
```

```{r}
#Table summarizing the average first vitals measurements by thirty day mortality.
mimic_icu_cohort %>% 
  group_by(thirty_day_mort) %>%
  summarise(mean_temp = mean(na.omit(temp_far)),
            mean_resp_rate = mean(na.omit(resp_rate)),
            mean_heart_rate = mean(na.omit(heart_rate)),
            mean_syst_bp = mean(na.omit(syst_bp)),
            mean_avg_bp = mean(na.omit(mean_bp))
  )
```
```{r}
#Table summarizing thirty day mortality by first ICU unit
mimic_icu_cohort %>% 
  group_by(first_careunit) %>%
  summarise(thirty_day_mortality = sum(na.omit(thirty_day_mort)))

ggplot(mimic_icu_cohort, aes(x = first_careunit, y = thirty_day_mort)) + 
  geom_col() + theme_minimal()
```

